Calendar-based graphics for visualizing people's daily schedules

### 2017-07-26 main.Rmd
rewrite scaling paragraph

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Facetting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides more details of the temporal patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

@@

(ref:flinders-2016-cap) The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a 3 by 4 monthly grid represents all the traffic in 2016. The disparities between weekday and weekend along with public holiday are immediately apparent.

@@

(ref:year-diagram-cap) Illustration of the indexing layout for months of one year.

@@

Reference lines dividing each cell and block as well as labels indicating weekday and month are also provided in order to make calendar-based graphics more accessible and informative.

@@

The algorithm can be extended to display data from a single month up to a few years. The number of rows and columns in the layout can be specified with the arguments \code{nrow} and \code{ncol}. If the one would like to visualize data spanning over three years, for example, \code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid out horizontally. The vertical direction can be enabled by swapping $i$ and $j$ in equation \ref{eq:grid}, which occurs when the argument \code{dir} is set to \code{"v"}. This benefits users who are accustomed to calendars that are organized vertically, which is common in some countries. Next we shall describe some of the arguments that allow for alternate displays.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} is considered as the spiral display of Figure \ref{fig:flinders-2016}.

@@

(ref:flinders-polar-cap) Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates. The periodic behaviors on workdays are clearly visible.

@@

Grouping the cells based on the same time period gives rise to the different scales. For example, the minimums and maximums obtained from each cell produce a local scale. The overall variation gives way to the individual shape. Figure \ref{fig:flinders-free} is an example of plotting line charts scaled locally. The daily variation is more distinctive compared to Figure \ref{fig:flinders-2016}.

Similarly, the same $j$ indexing day of the week is grouped to compute the scales in days of a week; in other words, the same value within a given $j$ corresponds to the same position across each $j$\textsuperscript{th} cell. To construct the scales for days of the month, $g$ representing day of the month is used. This makes it easier to compare shape of a given day across each month block. The scaling option combined with the calendar layouts offers a number of varieties to construct the plot.

@@

(ref:boxplot-cap) Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing loess smooth line for each day. It shows the hourly distribution in the city as a whole. Christmas stands out as the quietest day in December.

@@

(ref:chn-cap) Figure \ref{fig:boxplot} with Chinese text labels.

@@

The calendar-based visualization provides data plots in the familiar (at least for the Western world) format of an everyday tool. Special events for the region, like Anzac Day in Australia, or Thanksgiving Day in the USA, more easily pop out to the viewer as public holidays, rather than a typical work day.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

We propose a method for organising and visualising temporal data, collected on sub-daily interval, into a calendar layout. We hope this calendar-based visualization to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides more details of the temporal patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

@@

(ref:flinders-2016-cap) The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a $3 \times 4$ monthly grid represents all the traffic in 2016. The disparities between weekday and weekend along with public holiday are immediately apparent.

@@

(ref:year-diagram-cap) Illustration of the indexing layout for months of one year, with $M = 3$ and $N = 4$.

@@

In order to make calendar-based graphics more accessible and informative, reference lines dividing each cell and block as well as labels indicating weekday and month are also computed before plotting construction.

@@

The algorithm can be extended to display data from a single month up to a few years. The number of rows and columns in the layout can be specified with the arguments \code{nrow} and \code{ncol}. If the one would like to visualize data spanning over three years, for example, \code{nrow = 3} and \code{ncol = 12} would be an appropriate choice when assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid out horizontally. The vertical direction can be enabled by swapping $i$ and $j$ in equation \eqref{eq:grid}, which occurs when the argument \code{dir} is set to \code{"v"}. This benefits users who are accustomed to calendars that are organized vertically, which is common in some countries. Next we shall describe some of the arguments that allow for alternate displays.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} is equivalent to Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

(ref:flinders-polar-cap) Figure \ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The periodic behaviors on workdays are clearly visible.

@@

Grouping the cells based on the same time period gives rise to the different scales. For example, the minimums and maximums obtained from each cell produce a local scale. The overall variation gives way to the individual shape. Figure \ref{fig:flinders-free} is an example of plotting line charts scaled locally using \code{scale = "free"}. The daily variation is more distinctive compared to Figure \ref{fig:flinders-2016}.

The \code{free_wday} suggests that the scales are common for a given week day across all weeks, but different for each week days within a week. In other words, the same value within a given Monday, for example, corresponds to the same position across all Mondays but the different position in other week days. Similarly, the \code{free_mday} suggests free scaling for any day within a given month. The scaling option combined with the calendar layouts offers a number of varieties to construct the plot.

@@

(ref:boxplot-cap) Side-by-side boxplot of hourly counts across all the 43 sensors in December of 2016, with the superimposing loess smooth line for each day. It shows the hourly distribution in the city as a whole. There is one sensor attracting a larger number of people on New Year's Eve than the rest.

These calendar layouts, especially the monthly calendar, become widely used in most countries around the world. We offer languages other than English for text labelling. Figure \ref{fig:chn} shows the same plot as Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

(ref:chn-cap) The same side-by-side boxplot as Figure \ref{fig:boxplot}, but with the month and weekday labels in Chinese. It demonstrates the natural support for languages other than English.

@@

The calendar-based visualization provides data plots in the familiar format of an everyday tool. Special events for the region, like Anzac Day in Australia, or Thanksgiving Day in the USA, more easily pop out to the viewer as public holidays, rather than a typical work day.

### 2017-07-26 main.tex
rewrite scaling paragraph

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Facetting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 sensors installed that
count pedestrians every hour across the downtown area up until the end
of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of the foot traffic
at 3 different sensors. Figure \ref{fig:facet-time} provides more
details of the temporal patterns by faceting on day of the week. The
sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

@@

\caption[The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a 3 by 4 monthly grid represents all the traffic in 2016.
The disparities between weekday and weekend along with public holiday
are immediately apparent.]{The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a 3 by 4 monthly grid represents all the traffic in 2016.
The disparities between weekday and weekend along with public holiday
are immediately apparent.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

@@

\caption[Illustration of the indexing layout for months of
one year.]{Illustration of the indexing layout for months of
one year.}\label{fig:year-diagram}
\end{figure}
\end{CodeChunk}

@@

The algorithm can be extended to display data from a single month up to
a few years. The number of rows and columns in the layout can be
specified with the arguments \code{nrow} and \code{ncol}. If the one
would like to visualize data spanning over three years, for example,
\code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when
assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid
out horizontally. The vertical direction can be enabled by swapping
\(i\) and \(j\) in equation \ref{eq:grid}, which occurs when the
argument \code{dir} is set to \code{"v"}. This benefits users who are
accustomed to calendars that are organized vertically, which is common
in some countries. Next we shall describe some of the arguments that
allow for alternate displays.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} is considered
as the spiral display of Figure \ref{fig:flinders-2016}.

@@

\caption[Star plots of hourly pedestrian counts at
Flinders Street Station, which are line charts placed in polar
coordinates. The periodic behaviors on workdays are clearly visible.]{Star plots of hourly pedestrian counts at
Flinders Street Station, which are line charts placed in polar
coordinates. The periodic behaviors on workdays are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

@@

Grouping the cells based on the same time period gives rise to the
different scales. For example, the minimums and maximums obtained from
each cell produce a local scale. The overall variation gives way to the
individual shape. Figure \ref{fig:flinders-free} is an example of
plotting line charts scaled locally. The daily variation is more
distinctive compared to Figure \ref{fig:flinders-2016}.

Similarly, the same \(j\) indexing day of the week is grouped to compute
the scales in days of a week; in other words, the same value within a
given \(j\) corresponds to the same position across each
\(j\)\textsuperscript{th} cell. To construct the scales for days of the
month, \(g\) representing day of the month is used. This makes it easier
to compare shape of a given day across each month block. The scaling
option combined with the calendar layouts offers a number of varieties
to construct the plot.

@@

\caption[Side-by-side boxplot of hourly counts across all the
43 sensors in December of 2016, with the superimposing loess smooth line
for each day. It shows the hourly distribution in the city as a whole.
Christmas stands out as the quietest day in December.]{Side-by-side boxplot of hourly counts across all the
43 sensors in December of 2016, with the superimposing loess smooth line
for each day. It shows the hourly distribution in the city as a whole.
Christmas stands out as the quietest day in December.}\label{fig:boxplot}
\end{figure}
\end{CodeChunk}

@@

\caption[Figure \ref{fig:boxplot} with Chinese text labels.]{Figure \ref{fig:boxplot} with Chinese text labels.}\label{fig:chn}
\end{figure}
\end{CodeChunk}

>>>>

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Faceting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

@@

We propose a method for organising and visualising temporal data,
collected on sub-daily interval, into a calendar layout. We hope this
calendar-based visualization to provide insights into people's daily
schedules relative to events such as work days, weekends, holidays, and
special events. This work was originally motivated by studying foot
traffic in the city of Melbourne \citep{ped}. There have been 43 sensors
installed that count pedestrians every hour across the downtown area up
until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can
shed light into understanding people's daily schedules, or assisting
administration and business planning. We start off with the conventional
time series plot to catch a glimpse of such data. A small multiples,
shown in Figure \ref{fig:time-series-plot}, gives an overall picture of
the foot traffic at 3 different sensors. Figure \ref{fig:facet-time}
provides more details of the temporal patterns by faceting on day of the
week. The sensor data, like many temporal datasets on human behaviors,
lends itself to a number of exploratory data visualization challenges:

@@

\caption[The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a \(3 \times 4\) monthly grid represents all the traffic
in 2016. The disparities between weekday and weekend along with public
holiday are immediately apparent.]{The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a \(3 \times 4\) monthly grid represents all the traffic
in 2016. The disparities between weekday and weekend along with public
holiday are immediately apparent.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

@@

\caption[Illustration of the indexing layout for months of
one year, with \(M = 3\) and \(N = 4\).]{Illustration of the indexing layout for months of
one year, with \(M = 3\) and \(N = 4\).}\label{fig:year-diagram}
\end{figure}
\end{CodeChunk}

@@

The algorithm can be extended to display data from a single month up to
a few years. The number of rows and columns in the layout can be
specified with the arguments \code{nrow} and \code{ncol}. If the one
would like to visualize data spanning over three years, for example,
\code{nrow = 3} and \code{ncol = 12} would be an appropriate choice when
assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid
out horizontally. The vertical direction can be enabled by swapping
\(i\) and \(j\) in equation \eqref{eq:grid}, which occurs when the
argument \code{dir} is set to \code{"v"}. This benefits users who are
accustomed to calendars that are organized vertically, which is common
in some countries. Next we shall describe some of the arguments that
allow for alternate displays.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} is equivalent
to Figure \ref{fig:flinders-2016} placed in polar coordinates.

@@

\caption[Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The periodic behaviors on
workdays are clearly visible.]{Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The periodic behaviors on
workdays are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

@@

Grouping the cells based on the same time period gives rise to the
different scales. For example, the minimums and maximums obtained from
each cell produce a local scale. The overall variation gives way to the
individual shape. Figure \ref{fig:flinders-free} is an example of
plotting line charts scaled locally using \code{scale = "free"}. The
daily variation is more distinctive compared to Figure
\ref{fig:flinders-2016}.

The \code{free_wday} suggests that the scales are common for a given
week day across all weeks, but different for each week days within a
week. In other words, the same value within a given Monday, for example,
corresponds to the same position across all Mondays but the different
position in other week days. Similarly, the \code{free_mday} suggests
free scaling for any day within a given month. The scaling option
combined with the calendar layouts offers a number of varieties to
construct the plot.

@@

\caption[Side-by-side boxplot of hourly counts across all the
43 sensors in December of 2016, with the superimposing loess smooth line
for each day. It shows the hourly distribution in the city as a whole.
There is one sensor attracting a larger number of people on New Year's
Eve than the rest.]{Side-by-side boxplot of hourly counts across all the
43 sensors in December of 2016, with the superimposing loess smooth line
for each day. It shows the hourly distribution in the city as a whole.
There is one sensor attracting a larger number of people on New Year's
Eve than the rest.}\label{fig:boxplot}
\end{figure}
\end{CodeChunk}

These calendar layouts, especially the monthly calendar, become widely
used in most countries around the world. We offer languages other than
English for text labelling. Figure \ref{fig:chn} shows the same plot as
Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

\caption[The same side-by-side boxplot as Figure \ref{fig:boxplot},
but with the month and weekday labels in Chinese. It demonstrates the
natural support for languages other than English.]{The same side-by-side boxplot as Figure \ref{fig:boxplot},
but with the month and weekday labels in Chinese. It demonstrates the
natural support for languages other than English.}\label{fig:chn}
\end{figure}
\end{CodeChunk}
